---
title: "Cost of Stay Analysis (OB Patients with Births Complicated By Hypertension, Wisconsin, 2020)"
author: "Erik Westlund"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(forcats)
library(ggplot2)
library(ggtext)
library(kableExtra)
library(readxl)

source("theme.R")

raw_data <- readxl::read_excel("data/hcup/Wisconsin/HCUPnet_Community_DX1CCSR_County_WI_hyp_2020.xlsx", sheet = 4)

glimpse(raw_data)

```

# RMarkdown

RMarkdown is a powerful tool for creating dynamic documents that combine code, text, and visualizations. In this example, we'll use RMarkdown to create a bar chart of the average hospital costs per stay for a single county in Wisconsin.

You can show the actual code that gets run alongside the output, which is a very powerful way to communicate your analysis and allow others to inspect the logic of your approach.

Similar technologies include:

- Jupyter Notebooks
- Quarto

## Goal

Our goal is to create a bar chart of the average hospital costs per stay for a single county in Wisconsin.

## Prepare Data

We'll start by filtering the data to only include the counties with data, while selecting the columns we need.

```{r filter_data}
data <- raw_data |> 
  rename(
    county = `County`,
    average_hospital_costs_per_stay = `Average Hospital Costs per Stay (in $)`
  ) |> 
  select(county, average_hospital_costs_per_stay) |>
  filter(average_hospital_costs_per_stay != "*") |>
  mutate(average_hospital_costs_per_stay = as.numeric(average_hospital_costs_per_stay)) 

data |> head()
```

## Visualize

We'll create a bar chart of the average hospital costs per stay for each county.

```{r single_county}

p <- ggplot(data, aes(x = county, y = average_hospital_costs_per_stay)) +
  geom_col()

p
```

## Immediate Problems

* We are going to want the axes flipped because you cannot read the county names
* We may want to sort by cost and  by county
* We may want to add a title and axis labels
* The gray background is distracting and the lot seems to lack color

## Flip the Axis And Make The Plot Larger

```{r flip_axis, fig.width=8, fig.height=12}
p <- p + coord_flip()
p
```

## Improve Title and Axis Labels

```{r improve_labels, fig.width=8, fig.height=12}
p <- p + 
  labs(
    title = "Average Hospital Costs per Stay by County",
    subtitle = "OB Patients with Births Complicated By Hypertension, Wisconsin, 2020",
    x = element_blank(),
    y = "Average Hospital Costs per Stay ($)"
  )
p
```

## Improve Aesthetics

I have included a file, `theme.R`, which contains a custom theme for our plots. Let's apply this theme to improve the aesthetics of our plot.

This will:

* Remove teh gray background, which can be distracting
* Apply Johns Hopkins-themed colors

```{r improve_aesthetics, fig.width=8, fig.height=12}
p <- p + theme_jhu_bar()
p
```

## More Improvements

There are other improvement we may wish to make, such as:

* Sorting the bars by cost
* Adding dollars values to the bars
* Changing the color of the State and US Totals


## Sorting Bars

In R/ggplot2, this is achieved by reordering the factor levels of the x-axis variable.

```{r sort_bars, fig.width=8, fig.height=12}
p <- p + 
  aes(x = fct_reorder(county, average_hospital_costs_per_stay)) +
  labs(
    x = element_blank(),
  )

p
```

## Further Enhancements

So far, this looks similar to what we see with Stata. However, we can do more with R/ggplot2.

Let's:

* Color the bars for US Total/State Total
* Color the labels
* Add annotations just for US/State totals

```{r add_dollar_values, fig.width=8, fig.height=12}
data <- data |>
  mutate(county_factor = fct_reorder(county, average_hospital_costs_per_stay))

most_expensive <- data |> filter(average_hospital_costs_per_stay == max(average_hospital_costs_per_stay)) |> pull(county)
least_expensive <- data |> filter(average_hospital_costs_per_stay == min(average_hospital_costs_per_stay)) |> pull(county)

label_counties <- c("US Total", "State Total", most_expensive, least_expensive)

text_offset <- max(data$average_hospital_costs_per_stay) * 0.025

p <- ggplot(data, aes(x = county_factor, 
                      y = average_hospital_costs_per_stay, 
                      fill = case_when(
                        county == "US Total" ~ "US Total",
                        county == "State Total" ~ "State Total",
                        TRUE ~ "Other"
                      ))) +
  geom_col() + 
  geom_text(data = data |> filter(county %in% label_counties), 
            aes(label = scales::dollar(average_hospital_costs_per_stay), 
                y = average_hospital_costs_per_stay + text_offset),
            hjust = 0,
            size = 4, 
            fontface = "bold") +
  coord_flip() +
  scale_fill_manual(
    values = c("US Total" = "#E69F00", "State Total" = "#E69F00", "Other" = colors$HopkinsBlue),
    guide = "none"
  ) +
  expand_limits(y = max(data$average_hospital_costs_per_stay) * 1.1) +
  labs(
    title = "Average Hospital Costs per Stay by County",
    subtitle = "OB Patients with Births Complicated By Hypertension, Wisconsin, 2020",
    x = NULL,
    y = "Average Hospital Costs per Stay ($)"
  ) +
  theme_jhu_bar() +
  theme(
    axis.text.y = element_text(
      size = 10,
      face = ifelse(levels(data$county_factor) %in% c("US Total", "State Total"), "bold", "plain")
    )
  )

p

ggsave("output/2_hcup_wisconsin_hypertension_average_hospital_costs_per_stay.png", p, width = 8, height = 12)

```


