---
title: "Cost of Stay Analysis (OB Patients with Births Complicated By Hypertension, 9 States)"
author: "Erik Westlund"
date: "2025-03-19"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

library(dplyr)
library(forcats)
library(ggplot2)
library(ggridges)
library(ggtext)
library(kableExtra)
library(purrr)
library(readxl)
library(stringr)

source("theme.R")

```

## Goal

Our goal here will be to visualize differences in the cost of stay for OB patients with births complicated by hypertension across these states:

* California
* Georgia
* Louisiana
* Michigan
* Mississippi
* Oklahoma
* South Dakota
* Utah
* Wisconsin

## Read In Data

We will start by reading in the data from our HCUP data files.

We will do this by  making a list of the file paths and then using `lapply` to read in the data from each file.

```{r read_data}

data_files <- list(
  ca = "data/hcup/California/HCUPnet_Community_DX1CCSR_County_CA_hyp complicating birth_2020.xlsx",
  ga = "data/hcup/Georgia/HCUPnet_Community_DX1CCSR_County_GA_hyp complicating birth_2020.xlsx",
  la = "data/hcup/Louisiana/HCUPnet_Community_DX1CCSR_County_LA_hyp complicating birth_2020.xlsx",
  mi = "data/hcup/Michigan/HCUPnet_Community_DX1CCSR_County_MI_hyp complicating birth_2020.xlsx",
  ms = "data/hcup/Mississippi/HCUPnet_Community_DX1CCSR_County_MS_hyp complicating birth_2020.xlsx",
  ok = "data/hcup/Oklahoma/HCUPnet_Community_DX1CCSR_County_OK_hyp_2020.xlsx",
  sd = "data/hcup/South Dakota/HCUPnet_Community_DX1CCSR_County_SD_hyp_2020.xlsx",
  wi = "data/hcup/Wisconsin/HCUPnet_Community_DX1CCSR_County_WI_hyp_2020.xlsx"
)

dfs <- lapply(data_files, readxl::read_excel, sheet = 4)

```

## Prepare Data

We are going to want to combine all of the data into a single data frame.

We will do this by using a combination of `imap` from the `purrr` R package and `bind_rows` with a custom function that renames the columns and filters out rows with missing data.

Note how we are able to use the functional programming capabilities of R to chain together a series of operations in a clear and concise manner. To achieve this in Stata is, to use a technical term, a nightmare.

```{r combine_data}

data <- imap(dfs, ~ .x |> 
  rename(
    county = `County`,
    average_hospital_costs_per_stay = `Average Hospital Costs per Stay (in $)`
  ) |> 
  select(county, average_hospital_costs_per_stay) |> 
  filter(average_hospital_costs_per_stay != "*") |> 
  mutate(
    average_hospital_costs_per_stay = as.numeric(average_hospital_costs_per_stay),
    state = str_to_upper(.y)  # Add state abbreviation using the list key
  )
) |> bind_rows()

```

## Summarize our data frame

```{r summarize_data}

summary_data <- data |> 
  group_by(state) |> 
  summarize(
    n = n(),
    mean_cost = mean(average_hospital_costs_per_stay, na.rm = TRUE),
    median_cost = median(average_hospital_costs_per_stay, na.rm = TRUE),
    min_cost = min(average_hospital_costs_per_stay, na.rm = TRUE),
    max_cost = max(average_hospital_costs_per_stay, na.rm = TRUE)
  )

summary_data |> kable()
```

## Bar Chart of Average Cost of Stay

This takes the average cost of stay, by county, and plots it.

```{r bar_chart}
text_offset <- max(data$average_hospital_costs_per_stay) * 0.05

p <- ggplot(summary_data, aes(x = fct_reorder(state, mean_cost), y = mean_cost)) +
  geom_bar(stat = "identity", fill = colors$HopkinsBlue) +
  coord_flip() +
  geom_text(aes(label = scales::dollar(mean_cost), y = mean_cost + text_offset), vjust = 0) +
  labs(
    title = "Average Cost of Stay for OB Patients\nWith Births Complicated By Hypertension",
    subtitle = "By State, Average of County Averages",
    x = element_blank(),
    y = "Average Cost of Stay (USD)"
  ) +
  theme_jhu_bar() +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))
            
p
```

## What's Wrong With This?

We are not actually plotting the average cost of stay.

Rather, we are plotting the average cost of stay for each county, and then taking the averages.

What might be more insightful is to compare the distributions of average costs, by county, across states.

To do this, we can use a ridge plot with the help of the `ggridges` package.

```{r ridge_plot, fig.height=8, fig.width=8}

rp <- ggplot(data, aes(x = average_hospital_costs_per_stay, y = state)) +
  stat_density_ridges(
    fill = colors$HopkinsBlue,
    quantile_lines = TRUE, 
    quantiles = c(0.5), 
    alpha = 0.7
  ) +
  labs(
    title = "Distribution of Average Cost of Stay By County\nFor OB Patients With Births Complicated By Hypertension",
    subtitle = "Median Shown as Solid Line",
    x = "Average Cost of Stay By County (USD)",
    y = element_blank()
  ) +
  theme_jhu()
rp

ggsave("output/3_ridge_plot.png", rp, width = 8, height = 8)


```
